---
title: "Codebook"
date : "1st December 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = normalizePath("~/R")) 
```



##Dataset
==================================================================
**Human Activity Recognition Using Smartphones Dataset**
**Version 1.0**


The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain. See 'features_info.txt' for more details. 

**For each record it is provided:**


- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

**The dataset includes the following files:**

- 'README.txt'

- 'features_info.txt': Shows information about the variables used on the feature vector.

- 'features.txt': List of all features.

- 'activity_labels.txt': Links the class labels with their activity name.

- 'train/X_train.txt': Training set.

- 'train/y_train.txt': Training labels.

- 'test/X_test.txt': Test set.

- 'test/y_test.txt': Test labels.

The following files are available for the train and test data. Their descriptions are equivalent. 

- 'train/subject_train.txt': Each row identifies the subject who performed the activity for each window sample. Its range is from 1 to 30. 

- 'train/Inertial Signals/total_acc_x_train.txt': The acceleration signal from the smartphone accelerometer X axis in standard gravity units 'g'. Every row shows a 128 element vector. The same description applies for the 'total_acc_x_train.txt' and 'total_acc_z_train.txt' files for the Y and Z axis. 

- 'train/Inertial Signals/body_acc_x_train.txt': The body acceleration signal obtained by subtracting the gravity from the total acceleration. 

- 'train/Inertial Signals/body_gyro_x_train.txt': The angular velocity vector measured by the gyroscope for each window sample. The units are radians/second. 

**Notes:** 

- Features are normalized and bounded within [-1,1].
- Each feature vector is a row on the text file.

For more information about this dataset contact: activityrecognition@smartlab.ws

## Variables and Description
==================================================================
### Subject


Variable        |   Description
----------------|-------------------
subject         |   Subject IDs of the 30 people ranging from 1 to 30



### Activity 

Variable        |   Description
----------------|-------------------
activity        |   Each subject performs six activities as listed below:
                |   1 WALKING
                |   2 WALKING_UPSTAIRS
                |   3 WALKING_DOWNSTAIRS
                |   4 SITTING
                |   5 STANDING
                |   6 LAYING
                
### Features

Description :

The features selected for this database come from the accelerometer and gyroscope 3-axial raw signals tAcc-XYZ and tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. Similarly, the acceleration signal was then separated into body and gravity acceleration signals (tBodyAcc-XYZ and tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern:  
'-XYZ' is used to denote 3-axial signals in the X, Y and Z directions.


**Measurements extracted for this project (79 out of the  total 561 feature vectors):**

- mean(): mean

- std(): standard deviation

1.  freqBodyAcc.mean.X                      
2.  freqBodyAcc.mean.Y          
3.  freqBodyAcc.mean.Z                      
4.  freqBodyAcc.meanFreq.X                  
5.  freqBodyAcc.meanFreq.Y      
6.  freqBodyAcc.meanFreq.Z                  
7.  freqBodyAcc.std.X           
8.  freqBodyAcc.std.Y                       
9.  freqBodyAcc.std.Z           
10. freqBodyAccJerk.mean.X      
11. freqBodyAccJerk.mean.Y                  
12. freqBodyAccJerk.mean.Z      
13. freqBodyAccJerk.meanFreq.X  
14. freqBodyAccJerk.meanFreq.Y              
15. freqBodyAccJerk.meanFreq.Z  
16. freqBodyAccJerk.std.X                   
17. freqBodyAccJerk.std.Y       
18. freqBodyAccJerk.std.Z                   
19. freqBodyAccMag.mean         
20. freqBodyAccMag.meanFreq     
21. freqBodyAccMag.std                      
22. freqBodyBodyAccJerkMag.mean             
23. freqBodyBodyAccJerkMag.meanFreq         
24. freqBodyBodyAccJerkMag.std  
25. freqBodyBodyGyroJerkMag.mean            
26. freqBodyBodyGyroJerkMag.meanFreq
27. freqBodyBodyGyroJerkMag.std 
28. freqBodyBodyGyroMag.mean    
29. freqBodyBodyGyroMag.meanFreq
30. freqBodyBodyGyroMag.std                 
31. freqBodyGyro.mean.X                     
32. freqBodyGyro.mean.Y         
33. freqBodyGyro.mean.Z                     
34. freqBodyGyro.meanFreq.X                 
35. freqBodyGyro.meanFreq.Y     
36. freqBodyGyro.meanFreq.Z                 
37. freqBodyGyro.std.X          
38. freqBodyGyro.std.Y                      
39. freqBodyGyro.std.Z          
40. timeBodyAcc.mean.X                     
41. timeBodyAcc.mean.Y              
42. timeBodyAcc.mean.Z	                  
43. timeBodyAcc.std.X           
44. timeBodyAcc.std.Y                     
45. timeBodyAcc.std.Z           
46. timeBodyAccJerk.mean.X                  
47. timeBodyAccJerk.mean.Y      
48. timeBodyAccJerk.mean.Z                  
49. timeBodyAccJerk.std.X       
50. timeBodyAccJerk.std.Y                   
51. timeBodyAccJerk.std.Z       
52. timeBodyAccJerkMag.mean                 
53. timeBodyAccJerkMag.std      
54. timeBodyAccMag.mean                     
55. timeBodyAccMag.std          
56. timeBodyGyro.mean.X                     
57. timeBodyGyro.mean.Y         
58. timeBodyGyro.mean.Z                     
59. timeBodyGyro.std.X          
60. timeBodyGyro.std.Y                      
61. timeBodyGyro.std.Z          
62. timeBodyGyroJerk.mean.X                 
63. timeBodyGyroJerk.mean.Y     
64. timeBodyGyroJerk.mean.Z                 
65. timeBodyGyroJerk.std.X      
66. timeBodyGyroJerk.std.Y                  
67. timeBodyGyroJerk.std.Z      
68. timeBodyGyroJerkMag.mean                
69. timeBodyGyroJerkMag.std     
70. timeBodyGyroMag.mean                    
71. timeBodyGyroMag.std         
72. timeGravityAcc.mean.X                   
73. timeGravityAcc.mean.Y       
74. timeGravityAcc.mean.Z                   
75. timeGravityAcc.std.X        
76. timeGravityAcc.std.Y                    
77. timeGravityAcc.std.Z        
78. timeGravityAccMag.mean                  
79. timeGravityAccMag.std             

##The transformations and work done on the code 
==================================================================
*Below is the code as per the run_analysis R script, along with steps and relevant explanation*



###1. Setting the working directory and loading relevant libraries

```{r}
setwd("~/R")
library(dplyr)
library(data.table)
```

###2. Reading the test and train files

```{r}
xtest <- fread("X_test.txt")
ytest <- fread("y_test.txt")
xtrain <- fread("X_train.txt")
ytrain <-  fread("y_train.txt")
subtest <- fread("subject_test.txt")
subtrain <- fread("subject_train.txt")
```
Using the str() function it can then be seen that the dimensions of the datasets are as follows:

Dataset       |   Structure
--------------|---------------------
xtest         |   2947 obs. of  561 variables
xtrain        |   7352 obs. of  561 variables
ytest         |   2947 obs. of  1 variable
ytrain        |   7352 obs. of  1 variable
subtest       |   2947 obs. of  1 variable
subtrain      |   7352 obs. of  1 variable

This makes it easier to visualize how the all the datasets should be merged

###3. Combining the test and train file into a single dataframe

```{r}
xytest <- cbind(subtest,ytest,xtest)
xytrain <- cbind(subtrain,ytrain,xtrain)
xy <- rbind(xytest,xytrain)
```

###4. Reading the features file and extracting mean and std measurements

```{r}
features <- fread("features.txt")
pattern <- "mean|std"
```
###5. Extracting the 2 columns from features separately, to get the index/numbering and the corresponding names

*featindex is for column numbers to extract mean/std measurements*
*featname is for the descriptive variable names*

```{r}
featindex <- features[grepl(pattern,features$V2),1]
featname <- features[grepl(pattern,features$V2),2]
```

###6. Converting the combined data.table into a dataframe as we want a vector

```{r}
xy <- as.data.frame(xy, stringsAsFactors = FALSE)
```

###7. Unlisting featindex and featname as we want to match it to the columns in xymeanstd

```{r}
featindex <- unlist(featindex)
featname <- unlist(featname)
xymeanstd <- xy[,c(1,2,featindex+2)]
```

###8. Reading the activity labels file and using subsetting to replace the 2nd column with descriptive activity names

```{r}
activitylabel <- fread("activity_labels.txt")
activitylabel <- as.data.frame(activitylabel, stringsAsFactors = FALSE)
xymeanstd$V1.1 <- activitylabel[xymeanstd$V1.1,2]
colnames(xymeanstd) <- c("subject","activity", featname)
```

###9. Using gsub to tidy up the colnames- so t and f is changed to time and freq,while unneccesary brackets and dashes are removed

```{r}
colnames(xymeanstd) <- gsub("-",".",colnames(xymeanstd))
colnames(xymeanstd) <- gsub("^t","time",colnames(xymeanstd))
colnames(xymeanstd) <- gsub("f","freq",colnames(xymeanstd))
colnames(xymeanstd) <- gsub("[()]","",colnames(xymeanstd))
```

###10. Using the summarise, across and group by functions in dplyr to obtain a new dataset 

```{r}
summarydf <- xymeanstd %>%
  group_by(subject,activity) %>%
  summarise(across(.cols = (3:79), .fns = mean))
```
  
###11. writing the dataframe to a text file
```{r}
write.table(summarydf, file="summary.txt", sep = "\t", row.names = FALSE)
```
